BOLD Parameter Estimation Using Particle Filters

### 2010-07-02 doc/thesis/background.tex
started cleaning up notation in particle filter
more work....

\section{Solving the BOLD Model}
Unlike Statistical Parametric Mapping, the techniques described in this
section are all attempts to learn some version of the BOLD model. 
There have been quite a few efforts to quantify the parameters of the
various BOLD models.  Although \cite{Buxton1998} and \cite{Friston2000}
both proposed physiologically reasonable values for the model parameters, 
\cite{Friston2002} was the first paper to calculate the parameters based 
on actual FMRI data. However, in that case, the voxels were chosen from
regions that were detected as active by the GLM. It is therefore possible
that the parameters are biased toward parameters that fit the linear model.

@@

\section{Hybrid Methods}
In \cite{Riera2004}, a maximum
likelihood method for innovation processes was used, as described by
\cite{Ozaki1994}. \cite{Ozaki1994} uses a similar construction to a 
Kalman filter, to break the time series into a series of innovations,
for which Maximum Likelihood was performed. While this is probably the most
likely solution to give the correct output, there are a few problems. First, every
step in parameter space requires a recalculation of all the state variables. With
two or three parameters this is fine, more than that, and calculations can become
intractable. Additionally, there is no way to ensure a global minimum is reached.
Although this is also true of the methods in \autoref{sec:Nonlinear Least Squares},
those methods contained random elements designed to overcome this issue.
[results?]

>>>>

\section{Solving the BOLD Model}
Unlike Statistical Parametric Mapping, the techniques described in this
section are all attempts to learn some version of the BOLD model. 
There have been quite a few efforts to quantify the parameters of the
various BOLD models.  Although \cite{Buxton1998} and \cite{Friston2000}
both proposed physiologically reasonable values for the model parameters, 
\cite{Friston2002b} was the first paper to calculate the parameters based 
on actual FMRI data. However, in that case, the voxels were chosen from
regions that were detected as active by the GLM. It is therefore possible
that the parameters are biased toward parameters that fit the linear model.

@@

\subsection{Hybrid Methods}
In \cite{Riera2004}, a maximum
likelihood method for innovation processes was used, as described by
\cite{Ozaki1994}. \cite{Ozaki1994} uses a similar construction to a 
Kalman filter, to break the time series into a series of innovations,
for which Maximum Likelihood was performed. While this is probably the most
likely solution to give the correct output, there are a few problems. First, every
step in parameter space requires a recalculation of all the state variables. With
two or three parameters this is fine, more than that, and calculations can become
intractable. Additionally, there is no way to ensure a global minimum is reached.
Although this is also true of the methods in \autoref{sec:Nonlinear Least Squares},
those methods contained random elements designed to overcome this issue.
[results?]

### 2010-07-02 doc/thesis/intro.tex
more work....

In real FMRI data, there is a significant nonlinearity in response; with short sequences
responding disproportionately strong (\cite{Birn2001}, \cite{Wager2005}, \cite{Deneux2006}).
This nonlinearity is accounted for in the Balloon model, although \cite{Deneux2006}
shows that if there will be large variance in the length of signals, 
modeling Neural Habituation may be necessary to fully capture the range of responses. 
Stimuli that last longer than 4 seconds 
tend to be more linear, which is why block designs are so well accounted for
by the General Linear Model (\cite{Birn2001}, \cite{Deneux2006}). For this 
paper we will use only a simple version of the Balloon model, described by
\autoref{eq:bold}, to keep the solution tractable.

Another interesting result of \cite{Deneux2006} was the sensitivity analysis.
There it was found that many of the parameters compensate for each other,
implying that exact inference of parameters may not be possible without
constraint.

@@

\begin{table}[t]
\centering
\begin{tabular}{|c || c | c | c |}
\hline 
Parameter  & \cite{Friston2002} & \cite{Johnston2008} & \cite{Vakorin2007} \\
\hline
$\epsilon$ &  $N(.54 , .1 ^2)$  & $.069 \pm .014 $ & (NC)\\
$\tau_s  $ &  $N(1.54, .25^2)$  & $4.98 \pm 1.07 $ & $2.2$\\
$\tau_f  $ &  $N(2.46, .25^2)$  & $8.31 \pm 1.51 $ & $.45$\\
$\tau_0  $ &  $N(.98 , .25^2)$  & $8.38 \pm 1.5  $ & $.94$\\
$\alpha  $ &  $N(.33 , .45^2)$  & $.189 \pm .004 $ & $.4$ (NC) \\
$E_0     $ &  $N(.34 , .1 ^2)$  & $.635 \pm .072 $ & $.6$ (NC) \\
$V_0     $ &  $.03$ (NC)        & $.0149 \pm .006$ & (NC)\\
\hline
\end{tabular}
\caption{Parameters found by various studies. (NC) indicates that the value
wasn't calculated. \cite{Vakorin2007} made use of the values from \cite{Friston2002}
where not explicitly stated}
\label{tab:Params} 
\end{table}

>>>>

In real FMRI data, there is a significant nonlinearity in response; with short sequences
responding disproportionately strong (\cite{Birn2001}, \cite{Wager2005}, \cite{Deneux2006}).
This nonlinearity is accounted for in the Balloon model, although \cite{Deneux2006}
shows that if there will be large variance in the length of signals, 
modeling Neural Habituation may be necessary to fully capture the range of responses. 
Stimuli that last longer than 4 seconds 
tend to be more linear, which is why block designs are so well accounted for
by the General Linear Model (\cite{Birn2001}, \cite{Deneux2006}). For this 
paper we will use only a simple version of the Balloon model, described by
\autoref{eq:bold}, to keep the solution tractable, and because it is the most
well studied.

Another interesting result of \cite{Deneux2006} was the sensitivity analysis.
There it was found that the parameters are far from perpendicular,
implying that exact inference of parameters may not be possible without
constraint. This could explain the extreme discrepencies in \autoref{tab:Params}.

@@

%note to self, friston2002b's parameters are from a picture

\begin{table}[t]
\centering
\begin{tabular}{|c || c | c | c |}
\hline 
Parameter  & \cite{Friston2002} & \cite{Johnston2008} & \cite{Vakorin2007} & \cite{Deneux2006}\\
\hline
$\epsilon$ &  $N(.54 , .1 ^2)$  & $.069 \pm .014 $ & (NC) & .89\\
$\tau_s  $ &  $N(1.54, .25^2)$  & $4.98 \pm 1.07 $ & $2.2$ & 2.04 \\
$\tau_f  $ &  $N(2.46, .25^2)$  & $8.31 \pm 1.51 $ & $.45$ & 5.26\\
$\tau_0  $ &  $N(.98 , .25^2)$  & $8.38 \pm 1.5  $ & $.94$ & .27\\
$\alpha  $ &  $N(.33 , .45^2)$  & $.189 \pm .004 $ & $.4$ (NC) & .63 \\
$E_0     $ &  $N(.34 , .1 ^2)$  & $.635 \pm .072 $ & $.6$ (NC) & .33\\
$V_0     $ &  $.03$ (NC)        & $.0149 \pm .006$ & (NC) & .16\\
\hline
\end{tabular}
\caption{Parameters found by various studies. (NC) indicates that the value
wasn't calculated. \cite{Vakorin2007} made use of the values from \cite{Friston2002}
where not explicitly stated}
\label{tab:Params} 
\end{table}

### 2010-07-02 doc/thesis/methods.tex
more work....

\chapter{Methods}
\section{Prior Distribution}
For the BOLD signal all the parameters have been studied and have relatively well
known mean and variance, so a prior is helpful. 
For strictly positive parameters (members of $x$) we used a gamma distribution,
whereas for parameters that could be negative, we used a Gaussian distribution. In
both cases standard deviations twice that found in previous studies were used.

\section{Model}
As originally written in \autoref{sec:BOLD Physiology} the state variables
for the BOLD model are as follows:
\begin{eqnarray}
\dot{s} &=& \epsilon u(t) - \frac{s}{\tau_s} - \frac{f - 1}{\tau_f} \\
\dot{f} &=& s\\
\dot{v} &=& \frac{1}{\tau_0}(f - v^\alpha)\\
\dot{q} &=& \frac{1}{\tau_0}(\frac{f(1-(1-E_0)^f)}{E_0} - \frac{q}{v^{1-1/\alpha}})
\end{eqnarray}
Because there are potentially

>>>>

\chapter{Methods}
Although the particle filter I used for this work is a standard Regularized
Particle filter, as described in \cite{Arulampalam2002a}, optimizing the 
particle filter for use with FMRI data is non-trivial.

\section{Prior Distribution}
For the BOLD model described in \autoref{sec:BOLD Physiology}, several
different studies have endeavored to calculate parameters. The results
of these studies may be found in \autoref{tab:Params}, and the methods 
used for each may be found in \autoref{sec:Prior Works}. Unfortunately,
\cite{Friston2000} only studied small regions; and most research with
reasonable speed (including \cite{Friston2002b}) used these results as 
the source for their priors. 
The one exception is \cite{Johnston2008}, which came to an extremely different
set of distributions. For a particle filter, the choice of a prior is
the single most important design choice. Consequently for this work,
I wanted to be conservative; which meant going with the accepted result,
\cite{Friston2000}. This constrains the usefulness of the model to
areas that fall within the prior distribution, yet will allow results
to be comparable to other works. There is a significant need for better
estimates of the physiological parameters; and, while physical experiments
may not be possible, it would not be unreasonable to do a study with
exhaustive simulated annealing or hill climbing tests for multiple
regions and multiple patients. The purpose for this work is to determine
the fitness of particle filters for this task; because with good 
priors it can be extremely fast. Ultimately this algorithm may become
more useful as more studies are done on BOLD model that don't require
priors.

Therefore, because \cite{Friston2000} represents the current best
knowledge about the BOLD parameters, it forms the basis
for the prior distributions in this paper.

\begin{table}[t]
\centering
\begin{tabular}{|c || c | c |}
\hline 
Parameter & Distribution & $\mu$ & $\sigma$ \\
\hline
$\tau_0$ & Gamma & .98 & .5  \\
$\alpha$ & Gamma & .33 & .045\\
$E_0$    & Gamma & .34 & .03  \\
$V_0$    & Gamma & .03 & .006 \\
$\tau_s$ & Gamma & 1.54  & .5 \\
$\tau_f$ & Gamma & 2.46  & .5 \\
$\epsilon$ & Gamma & .64 & .4 \\
\hline
\end{tabular}
\caption{}
\label{tab:Prior} 
\end{table}

Note that although the mean remains the same for all the 
parameters other than $\epsilon$, the standard deviation is set
much higher to account for the disagreement between studies
(\autoref{tab:Params}). 
Because all the parameters are taken to be strictly positive, and the
standard deviations are approaching the mean, I used a gamma distribution.
This prevents the Gaussian from placing parameters in the nonsensical 
territory of negative activation, or negative time constants.

\section{Model}
As originally written in \autoref{sec:BOLD Physiology} the state variables
for the BOLD model are as follows:
\begin{eqnarray}
\dot{s} &=& \epsilon u(t) - \frac{s}{\tau_s} - \frac{f - 1}{\tau_f} \\
\dot{f} &=& s\\
\dot{v} &=& \frac{1}{\tau_0}(f - v^\alpha)\\
\dot{q} &=& \frac{1}{\tau_0}(\frac{f(1-(1-E_0)^f)}{E_0} - \frac{q}{v^{1-1/\alpha}})
\end{eqnarray}
The original assumption regarding particle filter models (\autoref{sec:Particle Filter Model})
included noise in the update of $x$, however that is not included here.
The reason for the difference is that cloud of particles is, to some extent,
able to model that noise. It is common, however, to actually model that noise
in particle filters by adding a random value to each updated state variable. 
Because the purpose of this particle filter is to learn the underlying distribution
of the static parameters, rather than precisely model the time course of the 
in the dynamic parameters ($\{s,f,v,q\}$) this noise is left out. Because the
BOLD model is dissipative, when no stimuli are applied, all the particles 
will decay to ($\{0,1,1,1\}$) which they should. If evolution noise were added, the 
particles would then weight particles based on the results of that noise, rather
than on the quality of the static parameters. Typical particle filter uses
also use this state noise as an exploratory measure; however given the 
high dimensionality of the system, a vast number of particles would be 
necessary for this to come to fruition. The noise in the BOLD signal is such
that several particles all tend to be equally correct; the particle filter will
still leave variance in the solution; which turned out to be the case.

Thus, at each time step, the states were linearized, and each state variable,
$s$, $f$, $v$, $q$, were changed according their previous rate of change.
Because of the difficulty of solving a system of nonlinear equations, I 
did not use the typical Runge-Kutta 4/5 technique to optimize step sizes. 
The cost of missing a feature in these differential equations typically leads
to non-real numbers shortly down the road. Because the non-real numbers do 
not come until the solution has had the chance to update two or three more
times, taking long steps can result in catastrophic and un-recoverable errors. 
Typically a step size of .001 was used, after finding that even .01 can
at times lead to the state equations careening out of control.

\section{Resampling}
The algorithm for resampling is described in \autoref{sec:Particle Filter Resampling}.
The primary design decision for resampling is the regularizing
kernel. As mentioned previously, the Gaussian kernel is extremely convenient,
because it is so easy to sample from. As discussed in \autoref{sec:Particle Filter Resampling},
as long as resampling is kept as a last resort, some amount of over-smoothing
won't impair convergence. Therefore, for this work I chose a Gaussian kernel of
bandwidth equal to the original distribution's covariance. Obviously this will
apply a rather large amount of smoothing to the distribution; however, on average
resampling is only applied every 10 to 20 measurements, and because randomization
is being applied to model updates this gives the filter some mobility.

The question of how often to resample is complex. The benefit of resampling
is that particles that were minimally important can give much more 
granularity in the middle of the distribution. At the same time, resampling
right after a batch of outliers can result in an incorrect distribution.
Because the memory of particle weights is not particularly long, "good" particles
could be removed by resampling very often. Additionally, because regularized
resampling is being with a large bandwidth, I did not want the particle filter
to be prevented from converging by resampling too often. Therefore, only
when the effective sample size dropped below 25 for two measurement points
in a row did I resample. This seemed to give good results, and avoided 
temporary drops in the $N_{eff}$ caused by outliers.
